### 2020-12-22 latex/introduction.tex
Fixed overfull hbox errors
Final edit, she be complete!

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\textwidth,angle=270]{images/kiiking.jpg}
    \caption{Kiiking, which translates to ``swing", involves a person standing
        and squatting on a giant swing until they are rotating around the bar.
    Image taken from \cite{kiiking-img}.}
    \label{fig:kiiking}
\end{figure}

@@

The modern concept of virtual nonholonomic constraints was described by
\citet{nhvc_dynamic_walking} in 2015, though there are references to more
primitive versions going back as early as the the year 2000
\cite{vnhc_human_robot_cooperation}.
Virtual nonholonomic constraints have been of most notable use in
bipedal locomotion \cite{nhvc_incline_walking,output_nhvc_bipedal_control},
where they have shown marked improvements in the robustness of walking gaits
when compared to previous control technqiues.
They have also been used for error-reduction in time-delayed teleoperation
\cite{vnhc_time_delay_teleop} and in the field of human-robot interaction
\cite{psd_based_vnhc_redundant_manipulator,haptic_vnhc}.
All these research areas use different formulations of virtual nonholonomic constraints;
\citet{hybrid_zero_dynamics_bipedal_nhvcs} have derived the so-called
``constrained dynamics" for hybrid mechanical systems under virtual nonholonomic
constraints, but to the best of our knowledge no one has attempted to generalize
the concept to abstract mechanical systems.

@@

\section{Statement of Contributions}
Here are the contributions of this thesis.
\begin{itemize}[label={}]
   \item \textbf{Chapter \ref{ch:vnhcs}} The development of the framework of
      virtual nonholonomic constraints.
      This includes the definition of simply actuated mechanical systems, virtual
      nonholonomic constraints, regular constraints, and energy injection.
      Theorem \ref{thm:vnhc-regularity} yields a computational characterization
      of regularity, while Theorem \ref{thm:zero-dynamics} explicitly finds the
      constrained dynamics for a certain class of systems.
   \item \textbf{Chapter \ref{ch:vlp}} An application of virtual nonholonomic
      constraints to the variable-length pendulum, based on a pumping technique
      used by children on standing swings.
      The chapter culminates in Theorem \ref{thm:vlp-energy-stabilization},
      which guarantees that a certain class of VNHCs will inject energy into
      this system.
   \item \textbf{Chapter \ref{ch:acrobot}} An application of virtual
      nonholonomic constraints to the acrobot, based on the motion performed by
      human gymnasts.
      The chapter ends with Theorem \ref{thm:acrobot-energy-stabilization},
      which proves that the constraint we design does in fact inject energy into
      the acrobot.
\end{itemize}

>>>>

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.6\textwidth,angle=270]{images/kiiking.jpg}
    \caption{Kiiking, which translates to ``swing", involves a person standing
        and squatting on a giant swing until they are rotating around the bar.
    Image taken from \cite{kiiking-img}.}
    \label{fig:kiiking}
\end{figure}

@@

The modern concept of virtual nonholonomic constraints was described by
\citet{nhvc_dynamic_walking} in 2015, though there are references to more
primitive versions going back as early as the the year 2000
\cite{vnhc_human_robot_cooperation}.
Virtual nonholonomic constraints have been of most notable use in
bipedal locomotion \cite{nhvc_incline_walking,output_nhvc_bipedal_control},
where they have shown marked improvements in the robustness of walking gaits
when compared to previous control technqiues.
They have also been used for error-reduction in time-delayed teleoperation
\cite{vnhc_time_delay_teleop} and in the field of human-robot interaction
\cite{psd_based_vnhc_redundant_manipulator,haptic_vnhc}.
All these research areas use different formulations of virtual nonholonomic constraints;
\citet{hybrid_zero_dynamics_bipedal_nhvcs} have derived the 
zero dynamics for hybrid mechanical systems under virtual nonholonomic
constraints, but to the best of our knowledge no one has attempted to generalize
the concept to abstract mechanical systems.

@@

\section{Statement of Contributions}
Here are the contributions of this thesis.
\begin{itemize}[label={}]
   \item \textbf{Chapter \ref{ch:vnhcs}} The development of 
      virtual nonholonomic constraints.
      This includes the definition of simply actuated mechanical systems, virtual
      nonholonomic constraints, regular constraints, and energy injection.
      Theorem \ref{thm:vnhc-regularity} yields a computational characterization
      of regularity, while Theorem \ref{thm:zero-dynamics} explicitly finds the
      constrained dynamics for a certain class of systems.
   \item \textbf{Chapter \ref{ch:vlp}} An application of virtual nonholonomic
      constraints to the variable-length pendulum, based on a pumping technique
      used by children on standing swings.
      The chapter culminates in Theorem \ref{thm:vlp-energy-stabilization},
      which guarantees that a certain class of VNHCs will inject energy into
      this system.
   \item \textbf{Chapter \ref{ch:acrobot}} An application of virtual
      nonholonomic constraints to the acrobot, based on the motion performed by
      human gymnasts.
      The chapter ends with Theorem \ref{thm:acrobot-energy-stabilization},
      which proves that the constraint we design does in fact inject energy into
      the acrobot.
\end{itemize}

### 2020-12-22 latex/vlp.tex
Final edit, she be complete!

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The Cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
   .
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
   .
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
   .
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
   .
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we get the Hamiltonian ~\eqref{eqn:simply-actuated-hamiltonian} whose
dynamics ~\eqref{eqn:simply-actuated-dynamics} resolve to  
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q)
      , \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau
        . \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \R\) affecting the dynamics of
\(p_l\), acting collinearly with the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This makes the VLP into an underactuated mechanical system with degree of
underactuation one. 
It is also a useful assumption because it means \((\mathbf{q},\mathbf{p})\) 
are simply actuated coordinates, which allows us to apply the theory of VNHCs we
developed in Chapter \ref{ch:vnhcs}.

Let us define the VNHC \(l = L(q,p)\), by which we mean we are
actually defining the VNHC \(h(\mathbf{q},\mathbf{p}) = l - L(q,p) = 0\) of
order 1.
The VLP satisfies
\(\nabla_q \Minv(\mathbf{q}) = \Zmat{2 \times 2}\).
By Theorem \ref{thm:vnhc-regularity},
\(h(\mathbf{q},\mathbf{p})\) is a regular VNHC whenever 
\(L(q,p)\) is \(C^2\) because
\[
   dh_{\mathbf{q}} \Minv(\mathbf{q})B = 
   \begin{bmatrix}
      -\pdiff{L}{q} & 1
   \end{bmatrix}
   \begin{bmatrix}
      \frac{1}{ml^2}  & 0 \\
      0 & \frac{1}{m}
   \end{bmatrix} 
   \begin{bmatrix}
      0 \\ 
      1
   \end{bmatrix}
   = \frac{1}{m}
\]
is always full rank.
The constraint manifold \(\Gamma\) is diffeomorphic to 
\(\SxR\), which is parameterized by the unactuated
phase \((q,p)\).
By Theorem \ref{thm:zero-dynamics}, the constrained dynamics
are described entirely by \((\dot{q},\dot{p})\) with \(l\)
replaced by \(L(q,p)\):
\begin{equation}\label{eqn:vlp-hamiltonian}
   \begin{cases}
      \dot{q} = \frac{p}{m L^2} \\
      \dot{p} = -mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
Note that we suppress the function notation of \(L(q,p)\) for clarity.

@@

\begin{equation}\label{eqn:vlp-optimal-controller}
   L^\star(\theta):= \begin{cases}
      \overline{l} & \theta \in [-\frac{\pi}{2},0[ \cup [\frac{\pi}{2}, \pi[ \\
      \underline{l} & \theta \in [-pi, -\frac{\pi}{2}[ \cup [0,\frac{\pi}{2}[ 
      .
   \end{cases}
\end{equation}

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \Sone \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) for the VLP injects energy 
   on the constraint manifold \(\Gamma \simeq \SxR\) if there
   exists \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \Sone
      ,
   \end{equation}
   with the property that the inequality is strict except at 
   \(\theta \in \{0, \frac{\pi}{2}, \frac{\pi}, \frac{3\pi}{2}\}\).  
   If instead 
   \[
    \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \leq 0 
         \text{ }\forall \theta \in \Sone
         ,
   \]
   the VNHC dissipates energy on \(\Gamma\).
\end{thm}
\begin{proof}
Since a system that gains energy per Definition \ref{defn:energy-gain}
cannot have stable equilibria,
we will first show that, under the given assumptions, the origin of the
constrained dynamics~\eqref{eqn:vlp-hamiltonian} is a repeller. 
For that we consider the negative time system obtained
from~\eqref{eqn:vlp-hamiltonian} by reversing the sign of the vector field:
\begin{equation}\label{eqn:negative-time-vlp}
   \begin{cases}
      \dot{q} = -\frac{p}{m L^2} \\
      \dot{p} = mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
The state space of~\eqref{eqn:negative-time-vlp} is 
\(\SxR\), and the system has two equilibria,
\((q,p) = (0,0)\) and \((q,p) = (\pi,0)\). 
We need to show that the equilibrium \((q,p) = (0,0)\) is
asymptotically stable for~\eqref{eqn:negative-time-vlp}.

@@

We now prove that almost all solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} escape compact sets in finite
time, \ie, that for each compact subset \(K\) of 
\(\SxR\), and for almost every condition 
\( (q_0,p_0) \in K \), there exists \( T > 0\) such that
\( (q(t),p(t)) \notin K \) for all \( t > T \), where \( (q(t),p(t)) \)
denotes the solution of~\eqref{eqn:vlp-hamiltonian} with 
\( (q(0),p(0)) = (q_0,p_0)\). 
In what follows, we denote by \( x = (q,p) \) the state of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}. 
Also, we denote by \(\Pi^-\) the stable manifold of the equilibrium \((\pi,0)\)
for system~\eqref{eqn:vlp-hamiltonian}, defined as
\[
   \Pi^- := \left\{ x(0) \in \SxR \mid
   \lim_{t \to \infty} x(t) = (\pi, 0) \right\}.
\]
The linearization of system~\eqref{eqn:vlp-hamiltonian} at the
equilibrium \((\pi,0)\) has a system matrix given by
\begin{equation}\label{eqn:linearized-vlp}
   \begin{bmatrix}
      0 & -\frac{1}{mL(0)^2} \\
      -m g L(0) & 0 \\
   \end{bmatrix}
   ,
\end{equation}
whose spectrum is \(\{\pm \sqrt{{g}/{L(0)}}\}\).  
Since one eigenvalue of ~\eqref{eqn:linearized-vlp} is positive and one is
negative, the stable manifold theorem implies that the stable manifold 
\( \Pi^- \) is a one-dimensional immersed submanifold of
\(\SxR\), and thus is a set of measure zero
\cite{stable_manifold}.

Let \( K \subset \SxR\) be an arbitrary compact set and let 
\( x(0) \in K \backslash \left(\Pi^- \cup \left\{(0,0)\right\}\right)\).
Since \( x(0) \notin \Pi^- \), we have that
\begin{equation}\label{eqn:vlp-x-does-not-converge}
   x(t) \nrightarrow_{t \to \infty} (\pi,0).
\end{equation}
Suppose, by way of contradiction, that for each \(T > 0\) there exists 
\(t' > T\) such that \(x(t') \in K\).  
Letting
\[
   k := \max\limits_{x \in K} E_\text{avg}(x)
   ,
\] 
and \(E_k =\{ x \mid E_\text{avg}(x) \leq k\} \), we have that 
\( K \subset E_k \), and thus \(x(t') \in E_k\).
We have shown earlier that the function \( E_\text{avg} \) is nonincreasing
along solutions of~\eqref{eqn:negative-time-vlp}, which implies that
\( E_\text{avg}(x(t)) \) is nondecreasing for solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}, and therefore the half orbit \(x([0,t'])\)
is contained fully within \(E_k\).  
In particular, \(x(T) \in E_k\). Since this is true for each \( T >0 \), we
deduce that \(x(t) \in E_k\) for all \(t \in [0,\infty[\).  
Since \(E_k\) is compact, \(x(t)\) has a positive limit set in \(E_k\) which, by
the Poincar\'{e}-Bendixson theorem, is either an equilibrium or a closed orbit
\cite{poincare_bendixson}.  
We will show this positive limit set must be an equilibrium and thereby reach a
contradiction.

In order to rule out closed orbits, suppose that the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} have a periodic solution \( z(t)\). 
Since \( E_\text{avg}(z(t)) \) is nondecreasing and \(z(t)\)
is periodic, it must be that \( E_\text{avg}(z(t)) \) is constant, or
\(\dot E_\text{avg}(z(t)) \equiv 0\). 
We have shown earlier
that\footnote{We have shown it for the negative time
system~\eqref{eqn:negative-time-vlp}, but changing the sign of the
vector field does not change the set where 
\(\dot E_\text{avg} =0\) nor the invariant sets.}
\[
   \{x \mid \dot E_\text{avg}(x) = 0\} = \left\{(q,p) \mid q = 0\right\}
   \cup \left\{(q,p) \mid p =0\right\},
\]
and the only invariant subset of this set is the union of two disjoint
equilibria. 
Since the orbit \(z(\R)\) is an invariant set which is
connected, \(z(\R)\) must be an equilibrium and cannot be a nontrivial
closed orbit.

Returning to the Poincar\`e-Bendixson theorem, the positive limit set
of \( x(t)\) must be an equilibrium and,
by~\eqref{eqn:vlp-x-does-not-converge}, this equilibrium must be 
\( (0,0)\). 
Since we have chosen \(x(0) \neq (0,0)\), and since \( (q,p) = (0,0) \) is a
repeller, \( x(t) \) cannot converge to \( (0,0)\), which gives a contradiction.

@@

Let us define \((L^\star)^{-}(\theta)\) by swapping the order of 
\(\underline{l}\) and \(\overline{l}\) in \(L^\star(\theta)\): 
\[
   (L^\star)^-(\theta) := \begin{cases}
      \underline{l} & \theta \in [-\frac{\pi}{2},0[ \cup [\frac{\pi}{2}, \pi[ \\
      \overline{l} & \theta \in [-pi, -\frac{\pi}{2}[ \cup [0,\frac{\pi}{2}[ 
      .
   \end{cases}
\]
Since this function \textit{minimizes} the derivative of \(E_\text{avg}\) under
the restriction \(l \in [\underline{l},\overline{l}]\), one might predict that 
\((L^\star)^{-}(\theta)\) is the optimal energy dissipation strategy for the VLP.
This is, in fact, true. \citet{pumping_swing_standing_squatting} showed 
that squatting at the lowest point of a swing and standing at the highest
(instead of standing and squatting, respectively) produces the
time-optimal trajectory for \textit{stopping} a standing swing.

@@

\begin{figure}
   \centering
   \includegraphics[width=0.8\textwidth]{images/kiiking_orbits.png}
   \caption{The orbits of the time-optimal controller \(L^\star(\theta)\)
      (black), the differentiable controller \(L_T(\theta)\) (red),
      and the smooth VNHC \(L_\frac{\pi}{2}(\theta)\) (cyan) employed by a 25
      year old Estonian male on a 3m tall Kiiking swing. All the controllers are
      initialized at \(q(0) = \frac{\pi}{32}\), and they each steadily gain
      energy until they reach a rotation.}
   \label{fig:vlp-sim-orbit}
\end{figure}

@@

\begin{figure}
   \centering
   \includegraphics[width=0.8\textwidth]{images/E_kiiking_ltheta.png}
   \caption{The energy \(E_\text{avg}(q,p)\) of a 25 year old Estonian male on a
      3m tall Kiiking swing who employs the smooth VNHC
   \(L_\frac{\pi}{2}(\theta)\).}
   \label{fig:vlp-sim-energy}
\end{figure}

>>>>

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The Cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
   .
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
   .
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
   .
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
   .
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we get the Hamiltonian ~\eqref{eqn:simply-actuated-hamiltonian} whose
dynamics ~\eqref{eqn:simply-actuated-dynamics} resolve to  
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q)
      , \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} 
        ,\\
        \dot{l} = \frac{p_l}{m} 
        ,\\
        \dot{p} = -mgl\sin(q) 
        ,\\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau
        . \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \R\) affecting the dynamics of
\(p_l\), acting collinearly with the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This makes the VLP into an underactuated mechanical system with degree of
underactuation one. 
It is also a useful assumption because it means \((\mathbf{q},\mathbf{p})\) 
are simply actuated coordinates, which allows us to apply the theory of VNHCs we
developed in Chapter \ref{ch:vnhcs}.

Let us define the VNHC \(l = L(q,p)\), by which we mean we are
actually defining the VNHC \(h(\mathbf{q},\mathbf{p}) = l - L(q,p) = 0\) of
order 1.
The VLP satisfies
\(\nabla_q \Minv(\mathbf{q}) = \Zmat{2 \times 2}\).
By Theorem \ref{thm:vnhc-regularity},
\(h(\mathbf{q},\mathbf{p})\) is a regular VNHC whenever 
\(L(q,p)\) is \(C^2\) because
\[
   dh_{\mathbf{q}} \Minv(\mathbf{q})B = 
   \begin{bmatrix}
      -\pdiff{L}{q} & 1
   \end{bmatrix}
   \begin{bmatrix}
      \frac{1}{ml^2}  & 0 \\
      0 & \frac{1}{m}
   \end{bmatrix} 
   \begin{bmatrix}
      0 \\ 
      1
   \end{bmatrix}
   = \frac{1}{m}
   ,
\]
is always full rank.
The constraint manifold \(\Gamma\) is diffeomorphic to 
\(\SxR\), which is parameterized by the unactuated
phase \((q,p)\).
By Theorem \ref{thm:zero-dynamics}, the constrained dynamics
are described entirely by \((\dot{q},\dot{p})\) with \(l\)
replaced by \(L(q,p)\):
\begin{equation}\label{eqn:vlp-hamiltonian}
   \begin{cases}
      \dot{q} = \frac{p}{m L^2} 
      , \\
      \dot{p} = -mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
Note that we suppress the function notation of \(L(q,p)\) for clarity.

@@

\begin{equation}\label{eqn:vlp-optimal-controller}
   L^\star(\theta):= \begin{cases}
      \overline{l} & \theta \in [-\frac{\pi}{2},0[ \cup [\frac{\pi}{2}, \pi[ \\
      \underline{l} & \theta \in [-\pi, -\frac{\pi}{2}[ \cup [0,\frac{\pi}{2}[ 
      .
   \end{cases}
\end{equation}

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \Sone \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) for the VLP injects energy 
   on the constraint manifold \(\Gamma \simeq \SxR\) if there
   exists \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \Sone
      ,
   \end{equation}
   with the property that the inequality is strict except at 
   \(\theta \in \left\{0, \frac{\pi}{2}, \pi, \frac{3\pi}{2}\right\}\).  
   If instead 
   \[
    \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \leq 0 
         \text{ }\forall \theta \in \Sone
         ,
   \]
   the VNHC dissipates energy on \(\Gamma\).
\end{thm}
\begin{proof}
Since a system that gains energy per Definition \ref{defn:energy-gain}
cannot have stable equilibria,
we will first show that, under the given assumptions, the origin of the
constrained dynamics~\eqref{eqn:vlp-hamiltonian} is a repeller. 
For that we consider the negative time system obtained
from~\eqref{eqn:vlp-hamiltonian} by reversing the sign of the vector field:
\begin{equation}\label{eqn:negative-time-vlp}
   \begin{cases}
      \dot{q} = -\frac{p}{m L^2} 
      ,\\
      \dot{p} = mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
The state space of~\eqref{eqn:negative-time-vlp} is 
\(\SxR\), and the system has two equilibria,
\((q,p) = (0,0)\) and \((q,p) = (\pi,0)\). 
We need to show that the equilibrium \((q,p) = (0,0)\) is
asymptotically stable for~\eqref{eqn:negative-time-vlp}.

@@

We now prove that almost all solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} escape compact sets in finite
time, \ie, that for each compact subset \(K\) of 
\(\SxR\), and for almost every condition 
\( (q_0,p_0) \in K \), there exists \( T > 0\) such that
\( (q(t),p(t)) \notin K \) for all \( t > T \), where \( (q(t),p(t)) \)
denotes the solution of~\eqref{eqn:vlp-hamiltonian} with 
\( (q(0),p(0)) = (q_0,p_0)\). 
In what follows, we denote by \( x = (q,p) \) the state of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}. 
Also, we denote by \(\Pi^+\) the stable manifold of the equilibrium \((\pi,0)\)
for system~\eqref{eqn:vlp-hamiltonian}, defined as
\[
   \Pi^+ := \left\{ x(0) \in \SxR \mid
   \lim_{t \to \infty} x(t) = (\pi, 0) \right\}.
\]
The linearization of system~\eqref{eqn:vlp-hamiltonian} at the
equilibrium \((\pi,0)\) has a system matrix given by
\begin{equation}\label{eqn:linearized-vlp}
   \begin{bmatrix}
      0 & -\frac{1}{mL(0)^2} \\
      -m g L(0) & 0 \\
   \end{bmatrix}
   ,
\end{equation}
whose spectrum is \(\{\pm \sqrt{{g}/{L(0)}}\}\).  
Since one eigenvalue of ~\eqref{eqn:linearized-vlp} is positive and one is
negative, the stable manifold theorem implies that the stable manifold 
\( \Pi^+ \) is a one-dimensional immersed submanifold of
\(\SxR\), and thus is a set of measure zero
\cite{stable_manifold}.

Let \( K \subset \SxR\) be an arbitrary compact set and let 
\( x(0) \in K \backslash \left(\Pi^+ \cup \left\{(0,0)\right\}\right)\).
Since \( x(0) \notin \Pi^+ \), we have that
\begin{equation}\label{eqn:vlp-x-does-not-converge}
   x(t) \nrightarrow_{t \to \infty} (\pi,0).
\end{equation}
Suppose, by way of contradiction, that for each \(T > 0\) there exists 
\(t' > T\) such that \(x(t') \in K\).  
Letting
\[
   k := \max\limits_{x \in K} E_\text{avg}(x)
   ,
\] 
and \(E_k =\{ x \mid E_\text{avg}(x) \leq k\} \), we have that 
\( K \subset E_k \), and thus \(x(t') \in E_k\).
We have shown earlier that the function \( E_\text{avg} \) is nonincreasing
along solutions of~\eqref{eqn:negative-time-vlp}, which implies that
\( E_\text{avg}(x(t)) \) is nondecreasing for solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}, and therefore the half orbit \(x([0,t'])\)
is contained fully within \(E_k\).  
In particular, \(x(T) \in E_k\). Since this is true for each \( T >0 \), we
deduce that \(x(t) \in E_k\) for all \(t \in [0,\infty[\).  
Since \(E_k\) is compact, \(x(t)\) has a positive limit set in \(E_k\) which, by
the Poincar\'{e}-Bendixson theorem, is either an equilibrium or a closed orbit
\cite{poincare_bendixson}.  
We will show this positive limit set must be an equilibrium and thereby reach a
contradiction.

In order to rule out closed orbits, suppose that the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} have a periodic solution \( z(t)\). 
Since \( E_\text{avg}(z(t)) \) is nondecreasing and \(z(t)\)
is periodic, it must be that \( E_\text{avg}(z(t)) \) is constant, or
\(\dot E_\text{avg}(z(t)) \equiv 0\). 
We have shown earlier
that\footnote{We have shown it for the negative time
system~\eqref{eqn:negative-time-vlp}, but changing the sign of the
vector field does not change the set where 
\(\dot E_\text{avg} =0\) nor the invariant sets.}
\[
   \left\{x \mid \dot E_\text{avg}(x) = 0\right\} = \left\{(q,p) \mid q = 0\right\}
   \cup \left\{(q,p) \mid p =0\right\},
\]
and the only invariant subset of this set is the union of two disjoint
equilibria. 
Since the orbit \(z(\R)\) is an invariant set which is
connected, \(z(\R)\) must be an equilibrium and cannot be a nontrivial
closed orbit.

Returning to the Poincar\'e-Bendixson theorem, the positive limit set
of \( x(t)\) must be an equilibrium and,
by~\eqref{eqn:vlp-x-does-not-converge}, this equilibrium must be 
\( (0,0)\). 
Since we have chosen \(x(0) \neq (0,0)\), and since \( (q,p) = (0,0) \) is a
repeller, \( x(t) \) cannot converge to \( (0,0)\), which gives a contradiction.

@@

Let us define \((L^\star)^{-}(\theta)\) by swapping the order of 
\(\underline{l}\) and \(\overline{l}\) in \(L^\star(\theta)\): 
\[
   (L^\star)^-(\theta) := \begin{cases}
      \underline{l} & \theta \in [-\frac{\pi}{2},0[ \cup [\frac{\pi}{2}, \pi[ \\
      \overline{l} & \theta \in [-\pi, -\frac{\pi}{2}[ \cup [0,\frac{\pi}{2}[ 
      .
   \end{cases}
\]
Since this function \textit{minimizes} the derivative of \(E_\text{avg}\) under
the restriction \(l \in [\underline{l},\overline{l}]\), one might predict that 
\((L^\star)^{-}(\theta)\) is the optimal energy dissipation strategy for the VLP.
This is, in fact, true. \citet{pumping_swing_standing_squatting} showed 
that squatting at the lowest point of a swing and standing at the highest
(instead of standing and squatting, respectively) produces the
time-optimal trajectory for \textit{stopping} a standing swing.

@@

\begin{figure}
   \centering
   \includegraphics[width=0.8\textwidth]{images/kiiking_orbits.png}
   \caption{The orbits of the time-optimal controller \(L^\star(\theta)\)
      (black), the differentiable controller \(L_T(\theta)\) (magenta),
      and the smooth VNHC \(L_\frac{\pi}{2}(\theta)\) (cyan) employed by a 25
      year old Estonian male on a 3m tall Kiiking swing. All the controllers are
      initialized at \(q(0) = \frac{\pi}{32}\), and they each steadily gain
      energy until they reach a rotation.}
   \label{fig:vlp-sim-orbit}
\end{figure}

@@

\begin{figure}
   \centering
   \includegraphics[width=0.9\textwidth]{images/E_kiiking_ltheta.png}
   \caption{The energy \(E_\text{avg}(q,p)\) of a 25 year old Estonian male on a
      3m tall Kiiking swing who employs the smooth VNHC
   \(L_\frac{\pi}{2}(\theta)\).}
   \label{fig:vlp-sim-energy}
\end{figure}
